Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies

### 2022-09-05 paper.qmd
updates to draft

---
title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
author: "Dean Marchiori"
format: pdf
abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and diverse clinical phenotype. A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals, driven in part to the paucity of analysis-ready data. This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis as a dimension reduction technique, which is then subject to heirarchical clustering and interpretation. To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data. The analysis identified the classic phenotype for this condition, as well as two other distinct feature clusters. The first relates to findings of vocal cord paralysis and swallowing dysfunction, the other relates to more complex brain anomalies. Furthermore, the survival probability for those affected was found to decline sharply in early infancy, but exhibited a wide range of outcomes provisionally associated with variant type. However, interpretation of these results are guarded based on very low sample sizes. Ultra rare conditions present a challenge for clinicians and families, with often limited information and diverse possible outcomes. Despite challenges with sparse and inconsistent data, this analysis helps describe associations and trajectories of indivudals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
editor: visual
bibliography: refs.bib
execute: 
  echo: false 
---

@@

A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information). Presentations were of microcephaly, simplified gyration, hypomyelination, thin corpus callosum, mild cerebellar hypoplasia, brainstem hypoplasia, congenital arthrogryposis, diabetes mellitus, heart disease, severe encephalopathy and respiratory problems often leading to early demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.

@@

A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem. Notably, there is also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.

@@

WIn terms of patient survival, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival]. These methods in particular provide a mechanism for dealing with right-censoring, a common feature of clinical studies where the clinical outcome of interest may not be known by the end of the study period. Within the context of examining a single condition, @Crowe810 investigated comorbidity phenotypes and mortality risk in ischaemic heart disease patients. Here latent class analysis was performed to identify a small number of clusters in the patients which could then be included in survival analysis techniques to better understand the comparative outcomes of patients of this condition.

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in SMPD4. With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood. Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome. While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.

@@

This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better describe and understand this condition. This research will assist clinicians understand the typical and non-typical presentations of this challenging new condition. In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.

@@

The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires:

1.  Each variable (clinical phenotype) forms a column.
2.  Each observation (individual) forms a row.
3.  Each type of observational unit forms a table (every cell has just one item)

@@

| id                     |      short palpebral fissures       | large ears | simple helices | receding forehead | ... |
|------------------------|-----------------|-------|--------|-------|-------|
| Family 1- Individual 1 | 1                        | 1          | 0              | ...               |     |
| Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |

@@

Some data type conversion and categorical level standardisation was performed to ensure the data were in consistent and appropriate data types. For example, 'Gender' was not consistently coded, and 'Birth Weight' was encoded as a text string rather than a more useful numeric format. (@tbl-data-types).

@@

The final dataset consisted of 21 observations (one per individual) and 152 variables (one per clinical feature). In some cases, were obvious, these variables were left as qualitative character descriptions, in other cases as noted above, distinct lists of features were tranformed into binary indicators. Other fields include background information on the patient and various biomarkers. A full list of this data set is available in @tbl-dict. These data sets were packaged into an R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.

@@

Dimension reduction on this wide data set was performed using Multiple Correspondence Analysis (MCA) [@le2010multiple] in order to arrive at a set of meaningfully small dimensions that account for most of the variation in the data. Multiple Correspondence Analysis (MCA) is an analogy to Principle Component Analysis (PCA) for categorical data.

@@

@@

A sensible range of clusters were trialed ranging from 2 to 7. Each of these settings were trialled across a range of clustering
algorithms including AGNES hierarchical agglomerative [@kaufman2009finding], Kmeans [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self].

The results for each method were evaluated using three internal cluster validation measures, Connectivity [@handl2005computational], Sihouette width [@rousseeuw1987silhouettes] and Dunn Index [@dunn1974well].

The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics
and a sensible partioning of the data for this analysis.

@@

To model survival probability for all subjects in the combined data, the data are subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric indicator which is equal to 1 if the subject is deceased and 0 otherwise.

The Kaplan-Meier estimator [@kaplan1958nonparametric] is used to estimate the survival function of the data.

@@

The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book].

Next the Kaplan-Meier estimator was calculated, \[ON WHAT?\] stratified by the reported type of Single nucleotide polymorphisms (SNP) reported in the literature. A test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* package.

@@

The initial results of the MCA were inspected with a bi-plot and identified two individuals (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse. These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and wer subject to a missense variation in SMPD4. Further analysis of this influence is conducted below. For now, these subjects were removed to allow for more meaningful analysis.

@@

```{r fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

mca$eig |> 
  tibble::as_tibble(rownames = 'dimension') |> 
  gt() |> 
  fmt_number(columns = 2,
             decimals = 2) |> 
  fmt_percent(columns = c(`percentage of variance`, `cumulative percentage of variance`),
              scale_values = F) |> 
   tab_header(
    title = "MCA Eigenvalue Table",
    subtitle = "Percentage variance explained by MCA for each dimension"
  )
```

The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr.

The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline. Also significant are seizure and developmental delay.

@@

The third dimension is characterised by complex non-typical features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features. These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.

```{r ctr-plot, fig.height=10, fig.width=10, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 20 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  labs(y = '')
```

```{r fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables"
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

A cluster relating to structual brain anomalies is articulated with conditions such as cerebellar hypoplasia and agenesis of corpus collosum, along with small for gestational age.

@@

@@

```{r, fig.width=8,  warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."

@@

```{r, fig.height=8, fig.width=8,  warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with cluster assignment indicated by colour."
cbind(cluster_data, clust = factor(cluster_assignments)) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  arrange(var, clust) |> 
  ggplot(aes(dim_1, dim_2, colour = clust)) +
  geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(aes(label = var), size = 2, max.overlaps = 20) +
  geom_point(aes(colour = clust), position = "jitter") +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       colour = 'cluster')
```

@@

```{r, warning=FALSE}
#| label: fig-surv
#| fig-cap: "Kaplan Meier survival curve"
survminer::ggsurvplot_df(x,
                         color = "black",
                         conf.int = TRUE,
                         conf.int.fill = "gray",
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(),
                         legend = "none") +
  labs(x = "time (days)")

@@

A comparitive boxplot of survival times by variant type shows significant outliers for missense+splice and missense variations. However, even within missense hymozygous variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as indeed noted by @magini2019loss.

```{r, warning=FALSE}
# KM by variant

# box plot of variant 
ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
  geom_boxplot()
```

The log rank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.

```{r}
# log rank
tar_load(logrank)
logrank
```

@@

Further distictions are possible with more specific structural brainanomalies exhibiting as distinct sources of variation in the clininal phenotype.

Many of the outlying features identifes are specific facial or anatomical dysmophims. While a noted and core feature is around course facial features, those with a milder course had clinical reports in the available data that articulated these dysmophims in much more detail.

On a similar note, the reporting of 'talipes' as acondition seems distict as a source of variation when appearing as an isolated feature without the typical microcephasly and hypomyelination. However, there was significant heterogenity in how feet contractures were reported in clinical case reports. A distinction was made between reported cases with rocker bottom feed or congenital vertical talus and those with clubfeet/talipes. In practice its not clear how these diagnoses were arived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literaure where possible.

The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves. This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child. While the number of clusters chosen all represent valid partition of the data, a larger number was selected to align the goals of the analysis which was to uncover distinct groups. A smaller number would have likely yeilded isolated outliers from the main body of the features. This still occurs, however more substructure is detected with the higher cluster number.

In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show much longer survival into the childhood and in some cases into the second decade of life. The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.

Despite NEDMABA exhibting a diverse and hetergeneous clinical phenotype, it is possible to identify meaningful substructures in this data through the use of dimension reduction techniques (MCA) and cluster analysis. This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay. Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD. In additional to typical MRI findings of symplified gyral pattern and hypomyelination, there exists clusters of structural anomalies of the corpus calllosum and cereballa hypoplasia.

The survival of indivuduals with this condition is critical given early demise in infancy and childhood is a noted feature. This work identifies the survival probabilites and shows a one year survival rate of 0.403 (0.252, 0.644). A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57. This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to overestimate the survival probability.

@@

```{r}
# Profile table 
cbind(cluster_data, clust = cluster_assignments) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  select(clust, var) |> 
  arrange(clust) |> 
  kable()
```

>>>>

---
title: "Statistical Analysis of Clinical Phenotype and Patient Survival in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies"
author: "Dean Marchiori"
format: pdf
abstract: "A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding the associations of the heterogenous features present in the clinical phenotype and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data in reported cases. This research aims to collate and standardise available case reports to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of sparse, multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To quantify more accurate survival probabilities, Kaplan-Meier estimation is formulated to account for censoring in the data. The analysis identified the classic phenotype for this condition, as well as two other distinct feature clusters. The first relates to findings of vocal cord paralysis and swallowing dysfunction, the other relates to more complex brain anomalies. Furthermore, the survival probability for those affected was found to decline sharply in early infancy, but exhibited a wide range of outcomes provisionally associated with variant type. However, interpretation of these results are guarded based on very low sample sizes. Despite challenges with sparse and inconsistent data, this analysis represents the first of its kind to help describe associations and trajectories of individuals with this condition and can provide clinicians and genetic counsellors with better information to aide in decision making and support for families."
editor: visual
bibliography: refs.bib
execute: 
  echo: false 
---

@@

A detailed study by @magini2019loss involved 12 unrelated familes with 32 individuals (21 with detailed clinical information). Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.

@@

A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem. Notably, there was also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.

@@

In terms of patient survival, methods in statistical survival analysis are well suited and commonly used in the field of genomics [@chen2014survival]. These methods in particular provide a mechanism for dealing with right-censoring, a common feature of clinical studies where the clinical outcome of interest may not be known by the end of the study period.

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in SMPD4. With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood. Furthermore, the survival time of affected children is not well understood despite early-demise featuring as a severe outcome in many cases. While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.

@@

This will be the first in-depth analysis of early studies of this novel variant, which aims to produce statistical findings to better understand this newly described condition. This research will assist clinicians understand the typical and non-typical presentations of this challenging new condition. In addition, genetic counselling of families with affected children will benefit from enhanced analysis on outcomes of existing reported cases.

@@

The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item).

@@

| id                     | short palpebral fissures | large ears | simple helices | receding forehead | ... |
|-----------------|-----------|-----------|-----------|-----------|-----------|
| Family 1- Individual 1 | 1                        | 1          | 1              | 0                 | ... |
| Family 1- Individual 4 | 1                        | 0          | 0              | 1                 | ... |

@@

Data type conversion and categorical level standardisation was performed to ensure the data were in consistent and appropriate data types. For example, 'Gender' was not consistently coded, and 'Birth Weight' was encoded as a text string rather than a more useful numeric format. (@tbl-data-types).

@@

The final dataset consisted of 28 observations (one per individual) and 152 variables (one per clinical feature). These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various biomarkers. A full list of this data set is available in @tbl-dict. These data sets were compiled into a publicly available R Package called SMPD4 [@smpd4-data] in order to allow for reproducibility and sharing.

@@

Dimension reduction on this wide data set was performed using Multiple Correspondence Analysis (MCA) (an analogy to Principle Component Analysis (PCA) for categorical data) [@le2010multiple] in order to arrive at a set of meaningfully small dimensions that account for most of the variation in the data.

@@

The coordinates on the first few principal axes were inspected for any outliers or unusual data.

@@

A range of clustering algorithms were comparied including AGNES hierarchical agglomerative [@kaufman2009finding], k-means [@hartigan1979algorithm] and self-organising maps (SOM) [@kohonen2012self]. A sensible range of clusters from 2-5 were trialed for each clustering algorithm. The results for each method were evaluated using three measures of stability: Average proportion of non-overlap (APN), Average Distance (AD) and Average Distance Between Means (ADM) [@datta2003comparisons]. Stability measures compare the full clustering result with a result based on dropping meach of the columns one at a time. These metrics were selected as we are representing each data point (clinical feature) with a number of MCA dimensions. A good cluster solution is one that is robust and meaningful across all of these dimensions [@fig-clvalid].

The optimal method and cluster number was selected based on a trade off between the best performing method using the evaluation metrics and a sensible partioning of the data for this analysis [@tbl-eig].

@@

To model survival probability for all subjects in the combined data, the data are subsetted to include *survival_time* which is either the number of days the individual survived for, or the age in days at last follow up, and *deceased* a numeric indicator which is equal to 1 if the subject is deceased and 0 otherwise. The Kaplan-Meier estimator [@kaplan1958nonparametric] is used to estimate the survival function of the data.

@@

The baseline estimator of all individuals was calculated using the *survival* package in R [@survival-book]. Next the Kaplan-Meier estimator was calculated, stratified by the reported type of genetic variation reported in the literature. A logrank test was performed to detect differences in the survival curves using methods from @harrington1982class, again implemented in the *survival* R package.

@@

The initial results of the MCA were inspected with a bi-plot and identified two subjects (17, 18) from the data that were clear outliers and sat significatly beyound the 95% confidence ellipse. These individuals represented two twins from @magini2019loss who were described with a much milder phenotype and were subject to a missense variation in *SMPD4*. Further analysis of this influence is conducted below. For now, these subjects were removed to allow for more meaningful analysis.

@@

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplotprelim
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables. A 95% confidence ellipse is plotted in red around the origin."
fviz_mca_biplot(
  mca_prelim,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point", "text"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.2, contrib = NULL),
  palette = "grey",
  col.ind = "red",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

@@

mca$eig |> 
  tibble::as_tibble(rownames = 'dimension') |> 
  gt() |> 
  fmt_number(columns = 2,
             decimals = 2) |> 
  fmt_percent(columns = c(`percentage of variance`, `cumulative percentage of variance`),
              scale_values = F)
```

The *contribution* of the categories were ranked within each of the first four MCA dimensions See @fig-ctr. The first dimension is dominated by highly specific facial dysmophisms such as epicanthal folds, short philtrum and low anteriour hairline.

@@

The third dimension is characterised by complex features such as patent ductus arteriosus, respiratory infections, small for gestational age and further facial features. These are contrasted by the fourth principal dimension which is typical for respiratory distress including stridor and hypercapnia.

```{r ctr-plot, fig.height=5, fig.width = 8, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  theme(legend.position = 'bottom') +
  labs(y = '')
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot
#| fig-cap: "A bi-plot of the first two principal axes with grey cicles indicating the individual's projected position and black triangles indicaating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(1, 2),
  geom = c("point", "text"),
  geom.ind = c("point"),
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)
```

```{r fig.height=5, fig.width=5,  warning=FALSE}
#| label: fig-biplot2
#| fig-cap: "A bi-plot of principal axes 3 and 4, with grey cicles indicating the individual's projected position and black triangles indicaating the variables. A 95% confidence ellipse is plotted in grey around the origin."
fviz_mca_biplot(
  mca,
  axes = c(3, 4),
  geom = c("point", "text"),
  geom.ind = "point",
  geom.var = c("point", "text"),
  repel = TRUE,
  label = "all",
  invisible = "none",
  habillage = "none",
  addEllipses = T,
  ellipse.level=0.95,
  select.var = list(name = NULL, cos2 = 0.4, contrib = NULL),
  palette = "grey",
  col.ind = "grey80",
  col.var = "black",
  arrows = c(FALSE, FALSE),
  map = "symmetric",
  title = ""
)

@@

@@

A full articulation of feature to cluster assignment is given in @tbl-profile.

@@

```{r, warning=FALSE}
#| label: fig-clvalid
#| fig-cap: "cluster validation analysis showing various metrics for cluster evaluation from 2 to 10 clusters. The most optimal number of clusters are selected by picking the lowest value from one or many of these compariable metrics."

@@

```{r, fig.height=6, fig.width=6,  warning=FALSE}
#| label: fig-clust
#| fig-cap: "A projection of features on the first two principal components with k-means cluster assignment indicated by colour."

d <- cbind(cluster_data, clust = factor(cluster_assignments)) |> 
  rownames_to_column(var = 'var') |> 
  as_tibble() |> 
  arrange(var, clust)

hull <- d %>%
  group_by(clust) |> 
  slice(chull(dim_1, dim_2))

  ggplot(d, aes(dim_1, dim_2, colour = clust)) +
  geom_polygon(data = hull, alpha = 0.2, aes(fill = clust)) +
  geom_vline(xintercept = 0, lty = 2, colour = "grey50") +
  geom_hline(yintercept = 0, lty = 2, colour = "grey50") +
  geom_label_repel(data = d |> group_by(clust) |> slice_sample(n=4), aes(label = var), size = 3, show.legend = F) +
  geom_point(aes(colour = clust), position = "jitter", show.legend = F) +
  labs(x = "Dimension 1",
       y = "Dimension 2",
       fill = 'cluster') +
    guides(color='none')
```

@@

```{r, warning=FALSE}
#| label: fig-surv
#| fig-cap: "Kaplan Meier curve with the black line representing the survival probability and with grey shading indicating the 95% confidence interval."
survminer::ggsurvplot_df(x,
                         color = "black",
                         conf.int = TRUE,
                         conf.int.fill = "gray",
                         risk.table = TRUE, # Add risk table
                         risk.table.col = "strata", # Change risk table color by groups
                         linetype = "strata", # Change line type by groups
                         surv.median.line = "hv", # Specify median survival
                         ggtheme = theme_bw(),
                         legend = "none") +
  labs(x = "time (days)")

@@

A comparitive boxplot of survival times by variant type shows significant outliers for missense+splice and missense variations. However, even within missense hymozygous variants the survival time varies greatly. Other missense variants, such as heterozygous missense with gene deletion do not show the same increased survival time as noted by @magini2019loss.

```{r,fig.width=7, warning=FALSE}
# KM by variant

# box plot of variant 
ggplot(survival_data, aes(survival_time, variant,fill = variant_type.x)) +
  geom_boxplot() +
  theme(legend.position = 'bottom') +
  labs(fill = 'Variant Type',
       x = 'Survival Time')
```

The logrank test was carried out to compare the suvival curves of the various gene variants $\chi^2 = 14.8$ $(p = 0.04)$ however due to the small sample size, several variants had expected counts less than 5 which would create a violation in assumptions for the test.

```{r}
# log rank
tar_load(logrank)
```

@@

Many of the outlying features identifes are specific facial or anatomical dysmophims. While a noted and core feature is around course facial features, those with a milder course had clinical reports in the available data that articulated these dysmophims in much more detail. On a similar note, the reporting of 'talipes' as acondition seems distict as a source of variation when appearing as an isolated feature without the typical microcephasly and hypomyelination. However, there was significant heterogenity in how feet contractures were reported in clinical case reports. A distinction was made between reported cases with rocker bottom feed or congenital vertical talus and those with clubfeet/talipes. In practice its not clear how these diagnoses were arived at, and given the non-specific nature of this feature, care needs to be taken in distinguishing between similar manifesting features in the literaure where possible.

The cluster analysis does not aim to group individuals that are alike, but rather the clinical features themselves and how these contrasting clusters account for variation in the data. This highlights common sources of variation on the clinical phenotype of the condition itself, rather than typical coincident features of any one child. While the number of clusters chosen all represent valid partition of the data, some metrics indicated an optimal selection of 2 clusters. This type of partition is useful when segregating outliers from the rest of the data and is valid in some application. In this study, a larger number was selected to align the goals of the analysis which was to uncover distinct groups of features.

In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show much longer survival into the childhood and in some cases into the second decade of life. This work identifies a one year survival rate of 0.403 (0.252, 0.644). A naive estimate, using just those number of deceased patients as proportion of the study population is 0.57. This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability. The current literature suggests the type of gene variant has an impact on surivival time, which seems to be supported in this work through a comparison of multiple survival surves, stratified by variation type, however the limited sample size here prevents other meaningful analysis such as logrank tests or cox proprtional hazard modelling which are likely underpowered when comparing multiple groups.

Despite NEDMABA exhibting a diverse and hetergeneous clinical phenotype, it is possible to identify meaningful substructures in this data through the use of dimension reduction techniques (MCA) and cluster analysis. This has isolated the classic NEDMABA phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyal pattern and developmental delay. Furthermore it has isolated additional distinct groupings of features that are exhibited in patients such as vocal cord paralysis with tracheostomy, swallowing dysfunction, tube feeding and GERD. Furthermore, we were able to quanitfy the expected baseline survival probabilities for children with this condition as well as quantifying the significant variation in these estimates. Additional work is required with the benefit of more cases, to better understand the differences in survival curves for other factors associated with genotype.

@@

```{r}
#| label: tbl-profile
#| tbl-cap: Assignment of clinical features with their cluster. These clusters represent similar features projected onto the first four MCA principal axes. Clusters that are on opposing sides of an axes contract with each other and account for the largest amount of variation in the data. 
cbind(cluster_data, cluster = cluster_assignments) |> 
  rownames_to_column(var = 'variable') |> 
  as_tibble() |> 
  select(cluster, variable) |> 
  arrange(cluster) |> 
  kable()
```
